Entropy stable reduced order modeling of nonlinear conservation laws

### 2019-09-06 docs/ESDG_ROM/main.tex
updating docs

\usepackage{amsmath,amssymb,amsfonts,amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\theoremstyle{lemma}
\newtheorem{lemma}{Lemma}
\newtheorem*{remark}{Remark}
\theoremstyle{theorem}
\newtheorem{theorem}{Theorem}
\theoremstyle{assumption}
\newtheorem{assumption}{Assumption}

@@

\begin{abstract}
Reduced order models of nonlinear conservation laws in fluid dynamics do not typically inherit stability properties of the full order model. We construct projection-based reduced models of nonlinear conservation laws based on flux differencing which retain accuracy while inheriting a semi-discrete entropy inequality independently of discretization resolutions and parameters. 
\end{abstract}
\end{frontmatter}

@@

For certain PDEs, ``structure-preserving'' ROMs provide robustness and stability by preserving energetic properties of the full system at the discrete level.  ROMs which retain either Lagrangian \cite{lall2003structure, carlberg2015preserving} or Hamiltonian structure \cite{benner2012interpolation, gugercin2012structure, peng2016symplectic, chaturantabut2016structure, gong2017structure, afkham2017structure, afkham2018structure} have been constructed by combining Galerkin projection with an appropriate formulation of the full order model, and similar energy-conserving ROMs have been constructed for the incompressible Navier-Stokes equations \cite{farhat2014dimensional, farhat2015structure}.  The construction of structure-preserving ROMs to nonlinear conservation laws in fluid flow, however, remains an open problem.  For example, Galerkin projection yields ROMs which become unstable as the number of reduced basis functions (modes) is increased \cite{bui2007goal, carlberg2013gnat}.  As a result, stabilized discretizations are often employed.  Petrov-Galerkin ROMs, which use an alternative test basis \cite{maday2002blackbox, rozza2007stability, bui2007goal, serre2012reliable, amsallem2012stabilization, rozza2013reduced, ballarin2015supremizer, carlberg2017galerkin}, are a popular alternative, as are additional residual-based stabilization or dissipation terms \cite{wang2012proper, kalashnikova2014stabilization, caiazzo2014numerical, balajewicz2016minimal}.  Such models improve robustness in practice, though they do not provide a theoretical foundation for stability.

The construction of stable ROMs for the compressible Navier-Stokes equations is complicated by their complex structure.  Thus, practitioners often develop structure-preserving entropy stable ROMs for simplifications of the underlying PDE (such as the time-dependent \textit{linearized} compressible Navier-Stokes equations \cite{barone2009stable, kalashnikova2010stability, amsallem2012stabilization, kalashnikova2014construction}) and extrapolate such models to the full nonlinear equations.  A promising alternative is to enforce physical conditions such as such as kinetic energy preservation, which are empirically related to stability \cite{maboudi2018conservative}.  These methods significantly improve the robustness of numerical methods in practice, and can guarantee a discrete entropy inequality for systems of nonlinear conservation laws which yield stable split formulations \cite{fisher2013discretely}.

The approach taken in this work differs in the treatment of nonlinear terms.  In \cite{maboudi2018conservative}, equations are rewritten such that the nonlinear terms involve only polynomial (quadratic and cubic) nonlinearities, which can be evaluated exactly using precomputed matrices.  In this work, we approximate nonlinear terms using hyper-reduction techniques based on empirically computed quadrature rules \cite{an2008optimizing, hernandez2017dimensional}.  The hyper-reduction approach incurs approximation error; however, it is allows for the generalization of the split forms in \cite{maboudi2018conservative} to nonlinearly entropy stable formulations.

%
%Hyper-reduction methods
%\begin{itemize}
%\item Interpolatory: DEIM and QR-DEIM \cite{chaturantabut2010nonlinear, drmac2016new}, gappy POD \cite{bui2004aerodynamic, carlberg2011efficient}
%\item Element-based approaches: ECSW \cite{farhat2014dimensional, farhat2015structure}, EQP \cite{ryu2015extensions, yano2019lp, patera2017lp}
%\item Empirical cubature: \cite{an2008optimizing, hernandez2017dimensional}.  
%\end{itemize}
%
%Structure-preserving ROMs: 
%\begin{itemize}
%\item Linear Hamlitonian \cite{beattie2009interpolatory} 
%\item Nonlinear Hamiltonian systems: Gugercin \cite{gugercin2012structure}, Carlberg \cite{carlberg2015preserving}, Chaturantabut \cite{chaturantabut2016structure}, Gong \cite{gong2017structure}, Afkham et al.\ \cite{afkham2017structure, afkham2018structure}
%\item Conservation through explicit constraints over sub-meshes \cite{carlberg2017structure, carlberg2018conservative}
%\end{itemize}

In this work, we directly constructing entropy stable ROMs for nonlinear conservation laws through a modified Galerkin projection of an appropriate full order model.  The approach taken in this paper combines techniques from entropy stable finite volume schemes \cite{tadmor1987numerical, tadmor2003entropy} and entropy stable summation-by-parts (SBP) discretizations \cite{fisher2013high, carpenter2014entropy, chen2017entropy, crean2018entropy, chan2017discretely, chan2019skew} to produce entropy stable reduced order models.

The paper is organized as follows.  Section~\ref{sec:1} discusses the entropy inequality associated with systems of nonlinear conservation laws.  Section~\ref{sec:2} introduces a full order model on 1D periodic domains based on entropy stable finite volume schemes.  Section~\ref{sec:3} describes how to construct an entropy stable reduced basis approximation, while Section~\ref{sec:4} discusses entropy stable hyper-reduction techniques to reduce costs associated with the evaluation of nonlinear terms.  Section~\ref{sec:6} describes how to extend the aforementioned approaches to non-periodic boundary conditions, and Section~\ref{sec:7} describes the extension to higher dimensions.  We conclude in Section~\ref{sec:8} with numerical experiments verifying theoretical results.

\section{Entropy inequalities for nonlinear conservation laws}
\label{sec:1}
Let $\Omega$ denote some domain with boundary $\partial \Omega$.  General nonlinear conservation laws are expressed as a system of partial differential equations (PDEs) 
\begin{equation}
\pd{\bm{u}}{t}  + \sum_{i=1}^d\pd{\bm{f}_i(\bm{u})}{x_i} = 0, \qquad 
S(\bm{u}) \text{ is a convex function}, \qquad
\bm{v}(\bm{u}) = \pd{S}{\bm{u}},
\label{eq:nonlineqs}
\end{equation}
where $\bm{u}\in \mathbb{R}^n$ are the conservative variables, $\bm{f}_j$ are nonlinear fluxes, and $\bm{v}(\bm{u})$ are the \textit{entropy variables}.  By multiplying (\ref{eq:nonlineqs}) by the entropy variables, viscosity solutions \cite{oleinik1957discontinuous, kruvzkov1970first}  of popular fluid systems (e.g., shallow water, compressible Euler and Navier-Stokes \cite{hughes1986new, chen2017entropy}) can be shown to satisfy 
\begin{equation}
\int_{\Omega}\pd{S(\bm{u})}{t}\diff{\bm{x}} + \sum_{i=1}^d \int_{\partial \Omega} \LRp{\bm{v}^T\bm{f}_i(\bm{u}) - \psi_i(\bm{u})}n_i \leq 0\label{eq:entropyineq},
\end{equation}
where $n_i$ denotes the $i$th component of the outward normal vector.  
The entropy inequality (\ref{eq:entropyineq}) is the analogue of energy stability for nonlinear conservation laws \cite{mock1980systems, harten1983symmetric}.  
However, due to the use of inexact treatment of nonlinear terms (e.g., from collocation approximations or inexact quadrature), most numerical methods for nonlinear conservation laws do not satisfy a discrete analogue of this stability condition.

@@

The reduced order models in this work are constructed using full order models constructed from discretely entropy stable finite volume methods \cite{tadmor1987numerical}.  For simplicity, we illustrate the construction of full and reduced order models on a periodic 1D domain.  We note that, while the reduced order model will be constructed from an entropy stable scheme, the solution snapshots do not need to be generated by entropy stable full order model.  Throughout this work, we will assume that the both full and reduced models yield ``physically relevant'' solutions, such that the entropy is a convex function.  For example, for the compressible Euler and Navier-Stokes equations, we will assume that the thermodynamic variables (density and energy/temperature/pressure) are positive.  Guaranteeing both positivity and accuracy of numerical solutions for general discretizations remains challenging \cite{guermond2016invariant, guermond2019invariant}, and will be explored in future work.

A key ingredient of the full and reduced order models considered in this work are \textit{entropy conservative} numerical fluxes.  Let $\bm{u}_L, \bm{u}_R$ denote left and right states.  Then, a two-point numerical flux $\bm{f}_S(\bm{u}_L, \bm{u}_R)$ is entropy conservative if it satisfies the following three conditions
\begin{gather}
\bm{f}_S(\bm{u},\bm{u}) = \bm{f}(\bm{u}), \qquad \text{(consistency)} \label{eq:esflux}\\
\bm{f}_S(\bm{u}_L,\bm{u}_R) = \bm{f}_S(\bm{u}_R,\bm{u}_R), \qquad \text{(symmetry)} \nonumber\\
\LRp{\bm{v}_L-\bm{v}_R}^T\bm{f}_S(\bm{u}_L,\bm{u}_R) = \psi(\bm{u}_L) - \psi(\bm{u}_R), \qquad \text{(entropy conservation)}\nonumber,
\end{gather}
Let the domain be decomposed into $K$ cells of size $\Delta x$, and let ${\LRp{\bm{u}_h}_i}$ denote the vector containing mean values of the conservation variables over the $i$th cell.  An entropy conservative finite volume method results from discretizing the integral form of the conservation law as follows
\begin{align}
&\td{{\LRp{\bm{u}_h}_1}}{t} + \frac{\bm{f}_S({\LRp{\bm{u}_h}_2},\LRp{\bm{u}_h}_1)-\bm{f}_S({\LRp{\bm{u}_h}_1},{\LRp{\bm{u}_h}_K})}{\Delta x} = \bm{0}, \label{eq:nonlineqs1d}\\
&\td{{\LRp{\bm{u}_h}_i}}{t} + \frac{\bm{f}_S({\LRp{\bm{u}_h}_{i+1}},\LRp{\bm{u}_h}_i)-\bm{f}_S({\LRp{\bm{u}_h}_i},{\LRp{\bm{u}_h}_i-1})}{\Delta x} = \bm{0}, \qquad 1 < i < K \nonumber\\
&\td{{\LRp{\bm{u}_h}_K}}{t} + \frac{\bm{f}_S({\LRp{\bm{u}_h}_1},\LRp{\bm{u}_h}_K)-\bm{f}_S({\LRp{\bm{u}_h}_K},{\LRp{\bm{u}_h}_K-1})}{\Delta x} = \bm{0},\nonumber
\end{align}
where periodicity is imposed through the equations for ${\LRp{\bm{u}_h}_1}, {\LRp{\bm{u}_h}_K}$.

@@

For a system of $n$ nonlinear conservation laws, all matrices are treated in a Kronecker product fashion.  Let the vector $\bm{u}_h$ now correspond to the concatenated vector of components of the solution $\bm{u}_h = [\bm{u}_h^1, \ldots, \bm{u}_h^n]$.  Then, the matrix formulation of (\ref{eq:nonlineqs1d}) is given by
\begin{gather*}
\Delta x\td{\bm{u}_h}{t} + \LRp{\LRp{\bm{I}_{n \times n} \otimes \bm{Q} }\circ \bm{F}}\bm{1} = \bm{0},\\
\bm{F} = \begin{bmatrix}
\bm{F}_{1} & &\\
& \ddots &\\
&& \bm{F}_{n} \\
\end{bmatrix}, \qquad 
\LRp{\bm{F}_{i}}_{jk} = \bm{f}_{i,S}\LRp{\LRp{\bm{u}_h}_j, \LRp{\bm{u}_h}_k}.
\label{eq:essys}
\end{gather*}
Here, $\bm{F}$ is a block diagonal matrix where the $i$th block corresponds to the evaluation of the $i$th component of the numerical flux $\bm{f}_{i,S}(\bm{u}_L,\bm{u}_R)$.  From this point on, we drop explicit references to the Kronecker product to simplify notation, so that (\ref{eq:es}) applies to both scalar nonlinear conservation laws and systems of nonlinear conservation laws.

@@

If $\bm{v}_h^T\bm{K}\bm{u}_h \geq 0$, then the solution satisfies a discrete dissipation of entropy
\begin{equation}
\Delta x \bm{1}^T\td{S(\bm{u}_h)}{t} = -\epsilon\bm{v}_h^T\bm{K}\bm{u}_h \leq 0.
\label{eq:entropybalancefom}
\end{equation}
The proof is given in \cite{tadmor2006entropy}, which we reproduce here.  We rewrite the viscous term as
\[
\bm{v}_h^T\bm{K}\bm{u}_h = \bm{v}_h^T \frac{1}{\Delta x^2}\begin{bmatrix}
\LRp{\bm{u}_h}_1 - \LRp{\bm{u}_h}_2\\ 
%\LRp{\LRp{\bm{u}_h}_2 - \LRp{\bm{u}_h}_1} - \LRp{\LRp{\bm{u}_h}_3 - \LRp{\bm{u}_h}_2}\\
\vdots\\
\LRp{\LRp{\bm{u}_h}_i - \LRp{\bm{u}_h}_{i-1}} - \LRp{\LRp{\bm{u}_h}_{i+1} - \LRp{\bm{u}_h}_{i}}\\
\vdots\\
\LRp{\bm{u}_h}_K- \LRp{\bm{u}_h}_{K-1} 
\end{bmatrix}.
\]
Since $\bm{v}_h$ and $\bm{u}_h$ are related through a differentiable invertible mapping, the mean value theorem implies that
\[
\LRp{\bm{u}_h}_i - \LRp{\bm{u}_h}_{i-1} = \LRp{\pd{\bm{u}}{\bm{v}}}_{i,i-1} \LRp{\LRp{\bm{v}_h}_i - \LRp{\bm{v}_h}_{i-1}}
\]
where $\LRp{\pd{\bm{u}}{\bm{v}}}_{i,i-1}$ is the positive-definite Jacobian matrix evaluated at some state between $(\bm{u}_h)_i$ and $(\bm{u}_h)_j$.  Substituting these expressions into the viscous term yields 
\begin{gather}
\bm{v}_h^T\bm{K}\bm{u}_h = 
%\bm{v}_h^T \frac{1}{\Delta x^2}\begin{bmatrix}
%\LRp{\pd{\bm{u}}{\bm{v}}}_{2,1} \LRp{\LRp{\bm{v}_h}_1 - \LRp{\bm{v}_h}_{2}}\\ 
%\vdots\\
%\LRp{\pd{\bm{u}}{\bm{v}}}_{i,i-1} \LRp{\LRp{\bm{v}_h}_i - \LRp{\bm{v}_h}_{i-1}} - \LRp{\pd{\bm{u}}{\bm{v}}}_{i+1,i} \LRp{\LRp{\bm{v}_h}_{i+1} - \LRp{\bm{v}_h}_{i}}\\
%\vdots\\
%\LRp{\pd{\bm{u}}{\bm{v}}}_{K,K-1} \LRp{\LRp{\bm{v}_h}_K - \LRp{\bm{v}_h}_{K-1}}
%\end{bmatrix}\\
 \frac{1}{\Delta x^2}\sum_{i=1}^{K-1} \LRp{\pd{\bm{u}}{\bm{v}}}_{i+1,i} \LRb{\LRp{\bm{v}_h}_{i} - \LRp{\bm{v}_h}_{i+1}}^2 \geq 0.
 \label{eq:entropydissfom}
\end{gather}

\section{An entropy conservative reduced order model}
\label{sec:3}
We can now formulate a reduced basis approximation on periodic domains using the entropy stable full order models described in the previous section.  Boundary conditions will be addressed later in this work.  We also note that this reduced order model is not practical, since the cost of evaluating nonlinear terms scales with the size of the full order model.  Later sections will discuss how to reduce the cost of nonlinear evaluations using appropriate hyper-reduction techniques.

@@

We will show that, while this formulation is not entropy conservative, a slight modification recovers semi-discrete entropy conservation.  We motivate this modification by first considering the time derivative term.  Because of the Galerkin projection step, we can no longer test with the vector of entropy variables.  However, we can test with an appropriate \textit{projection} of the entropy variables.  Let $\bm{v}_N$ denote coefficients of the projection of the entropy variables

@@

 \bm{v}\LRp{\bm{V}\bm{u}_N}^T \td{\bm{V}\bm{u}_N}{t} = \bm{1}^T\td{S(\bm{V}\bm{u}_N)}{t}.\nonumber

@@

Define the grid values of the projected entropy variables as $\tilde{\bm{v}}$ 

@@

\tilde{\bm{v}} = \bm{V}\bm{v}_N = \bm{V}\bm{V}^{\dagger} \bm{v}\LRp{\bm{V}\bm{u}_N}.

@@

It is possible to repeat the steps of the proof of Theorem~\ref{thm:ecfom} up to the point when we invoke the entropy conservation condition of (\ref{eq:esflux})

@@

We apply a two-step hyper-reduction procedure which preserves both the zero row sums and skew-symmetry of the hyper-reduced matrix $\bm{Q}_s$.  Rather than directly hyper-reducing the full order matrix $\bm{Q}$, we first construct a compressed ``modal'' matrix, then combine this with a projection operator based on the hyper-reduced points.

>>>>

\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{hhline}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\theoremstyle{lemma}
\newtheorem{lemma}{Lemma}
\newtheorem*{remark}{Remark}
\theoremstyle{theorem}
\newtheorem{theorem}{Theorem}
\theoremstyle{assumption}
\newtheorem{assumption}{Assumption}

@@

\begin{abstract}
Reduced order models of nonlinear conservation laws in fluid dynamics do not typically inherit stability properties of the full order model. We introduced projection-based reduced models of nonlinear conservation laws based on flux differencing which retain accuracy while inheriting a semi-discrete entropy inequality independently of the choice of basis, parameters, and discretization resolution. 
\end{abstract}
\end{frontmatter}

@@

For certain PDEs, ``structure-preserving'' ROMs provide robustness and stability by reproducing energetic properties of the full system at the discrete level.  ROMs which retain either Lagrangian \cite{lall2003structure, carlberg2015preserving} or Hamiltonian structure \cite{benner2012interpolation, gugercin2012structure, peng2016symplectic, chaturantabut2016structure, gong2017structure, afkham2017structure, afkham2018structure} have been constructed by combining Galerkin projection with an appropriate formulation of the full order model, and similar energy-conserving ROMs have been constructed for the incompressible Navier-Stokes equations \cite{farhat2014dimensional, farhat2015structure}.  The construction of structure-preserving ROMs to nonlinear conservation laws in fluid flow, however, remains an open problem.  For example, Galerkin projection yields ROMs which become unstable as the number of reduced basis functions (modes) is increased \cite{bui2007goal, carlberg2013gnat}.  As a result, stabilized discretizations are often employed.  Petrov-Galerkin ROMs, which use an alternative test basis \cite{maday2002blackbox, rozza2007stability, bui2007goal, serre2012reliable, amsallem2012stabilization, rozza2013reduced, ballarin2015supremizer, carlberg2017galerkin}, are a popular alternative, as are additional residual-based stabilization or dissipation terms \cite{wang2012proper, kalashnikova2014stabilization, caiazzo2014numerical, balajewicz2016minimal}.  Such models improve robustness in practice, though they do not provide a theoretical foundation for stability.

The construction of stable ROMs for the compressible Navier-Stokes equations is further complicated by their non-trivial structure.  Thus, practitioners often develop structure-preserving entropy stable ROMs for simplifications of the underlying PDE (such as the time-dependent \textit{linearized} compressible Navier-Stokes equations \cite{barone2009stable, kalashnikova2010stability, amsallem2012stabilization, kalashnikova2014construction}) and extrapolate such models to the full nonlinear equations.  A promising alternative is to enforce physical conditions such as such as kinetic energy preservation, which are empirically related to stability \cite{maboudi2018conservative}.  These methods significantly improve the robustness of numerical methods in practice, and can guarantee a discrete entropy inequality for systems of nonlinear conservation laws which yield stable split formulations \cite{fisher2013discretely}.

The approach taken in this work differs in the treatment of nonlinear terms.  In \cite{maboudi2018conservative}, equations are rewritten such that the nonlinear terms involve only polynomial (quadratic and cubic) nonlinearities, which can be evaluated exactly using precomputed matrices.  In this work, we approximate nonlinear terms using hyper-reduction techniques based on empirically computed quadrature rules \cite{an2008optimizing, hernandez2017dimensional}.  The hyper-reduction approach incurs approximation error and additional computational cost; however, it is allows for the generalization of the split forms in \cite{maboudi2018conservative} to nonlinearly entropy stable formulations.  We directly construct entropy stable ROMs for nonlinear conservation laws by combining hyper-reduction with a modified Galerkin projection of an appropriate full order model.  The approach taken in this paper combines techniques from entropy stable finite volume schemes \cite{tadmor1987numerical, tadmor2003entropy} and entropy stable summation-by-parts (SBP) discretizations \cite{fisher2013high, carpenter2014entropy, chen2017entropy, crean2018entropy, chan2017discretely, chan2019skew} to produce entropy stable reduced order models.

The paper is organized as follows.  Section~\ref{sec:1} introduces entropy inequalities associated with systems of nonlinear conservation laws.  Section~\ref{sec:2} introduces a full order model on 1D periodic domains based on entropy stable finite volume schemes.  Section~\ref{sec:3} describes how to construct an entropy stable reduced basis approximation, while Section~\ref{sec:4} discusses entropy stable hyper-reduction techniques to reduce costs associated with the evaluation of nonlinear terms.  Section~\ref{sec:6} describes how to extend the aforementioned approaches to non-periodic boundary conditions, and Section~\ref{sec:7} describes the extension to higher dimensions.  We conclude in Section~\ref{sec:8} with numerical experiments verifying theoretical results.

\section{Entropy inequalities for nonlinear conservation laws}
\label{sec:1}
Let $\Omega$ denote some domain with boundary $\partial \Omega$.  Nonlinear conservation laws are expressed as a system of partial differential equations (PDEs) 
\begin{equation}
\pd{\bm{u}}{t}  + \sum_{i=1}^d\pd{\bm{f}_i(\bm{u})}{x_i} = 0, \qquad 
S(\bm{u}) \text{ is a convex function}, \qquad
\bm{v}(\bm{u}) = \pd{S}{\bm{u}},
\label{eq:nonlineqs}
\end{equation}
where $\bm{u}\in \mathbb{R}^n$ are the conservative variables, $\bm{f}_i$ are nonlinear fluxes, and $\bm{v}(\bm{u})$ are the \textit{entropy variables}.  By multiplying (\ref{eq:nonlineqs}) by the entropy variables, viscosity solutions \cite{oleinik1957discontinuous, kruvzkov1970first}  of popular fluid systems (e.g., shallow water, compressible Euler and Navier-Stokes \cite{hughes1986new, chen2017entropy}) can be shown to satisfy 
\begin{equation}
\int_{\Omega}\pd{S(\bm{u})}{t}\diff{\bm{x}} + \sum_{i=1}^d \int_{\partial \Omega} \LRp{\bm{v}^T\bm{f}_i(\bm{u}) - \psi_i(\bm{u})}n_i \leq 0\label{eq:entropyineq},
\end{equation}
where $n_i$ denotes the $i$th component of the outward normal vector.  
The entropy inequality (\ref{eq:entropyineq}) is the analogue of energy stability for nonlinear conservation laws \cite{mock1980systems, harten1983symmetric}.  
However, due to the use of inexact treatment of nonlinear terms (e.g., from collocation approximations or inexact quadrature), most numerical methods for nonlinear conservation laws do not satisfy a discrete analogue of this stability condition.

@@

The reduced order models in this work are constructed using full order models based on entropy stable finite volume methods \cite{tadmor1987numerical}.  For simplicity, we illustrate the construction of full and reduced order models on a periodic 1D domain.  We note that, while the reduced order model will be constructed from an entropy stable scheme, the solution snapshots do not need to be generated by entropy stable full order model.  Throughout this work, we will assume that the both full and reduced models yield ``physically relevant'' solutions, such that the entropy is a convex function.  For example, for the compressible Euler and Navier-Stokes equations, we will assume that the thermodynamic variables (density and energy/temperature/pressure) are positive.  Guaranteeing both positivity and accuracy of numerical solutions for general discretizations remains challenging \cite{guermond2016invariant, guermond2019invariant}, and will be explored in future work.

A key ingredient of the full and reduced order models considered in this work are \textit{entropy conservative} finite volume numerical fluxes.  Let $\bm{u}_L, \bm{u}_R$ denote left and right states.  Then, a two-point numerical flux $\bm{f}_S(\bm{u}_L, \bm{u}_R)$ is entropy conservative if it satisfies the following three conditions
\begin{gather}
\bm{f}_S(\bm{u},\bm{u}) = \bm{f}(\bm{u}), \qquad \text{(consistency)} \label{eq:esflux}\\
\bm{f}_S(\bm{u}_L,\bm{u}_R) = \bm{f}_S(\bm{u}_R,\bm{u}_R), \qquad \text{(symmetry)} \nonumber\\
\LRp{\bm{v}_L-\bm{v}_R}^T\bm{f}_S(\bm{u}_L,\bm{u}_R) = \psi(\bm{u}_L) - \psi(\bm{u}_R), \qquad \text{(entropy conservation)}\nonumber,
\end{gather}
Let the domain be decomposed into $K$ cells of size $\Delta x$, and let ${\LRp{\bm{u}_h}_i}$ denote the vector containing mean values of the conservation variables over the $i$th cell.  An entropy conservative finite volume method results from discretizing the integral form of the conservation law as follows
\begin{align}
&\td{{\LRp{\bm{u}_h}_1}}{t} + \frac{\bm{f}_S({\LRp{\bm{u}_h}_2},\LRp{\bm{u}_h}_1)-\bm{f}_S({\LRp{\bm{u}_h}_1},{\LRp{\bm{u}_h}_K})}{\Delta x} = \bm{0}, \label{eq:nonlineqs1d}\\
&\td{{\LRp{\bm{u}_h}_i}}{t} + \frac{\bm{f}_S({\LRp{\bm{u}_h}_{i+1}},\LRp{\bm{u}_h}_i)-\bm{f}_S({\LRp{\bm{u}_h}_i},{\LRp{\bm{u}_h}_i-1})}{\Delta x} = \bm{0}, \qquad 1 < i < K \nonumber\\
&\td{{\LRp{\bm{u}_h}_K}}{t} + \frac{\bm{f}_S({\LRp{\bm{u}_h}_1},\LRp{\bm{u}_h}_K)-\bm{f}_S({\LRp{\bm{u}_h}_K},{\LRp{\bm{u}_h}_K-1})}{\Delta x} = \bm{0},\nonumber
\end{align}
where periodicity is imposed through the equations for ${\LRp{\bm{u}_h}_1}, {\LRp{\bm{u}_h}_K}$.

@@

For a system of $n$ nonlinear conservation laws, all matrices are treated in a Kronecker product fashion.  Let the vector $\bm{u}_h$ now correspond to the concatenated vector of components of the solution $\bm{u}_h = [\bm{u}_h^1, \ldots, \bm{u}_h^n]^T$.  Then, the matrix formulation of (\ref{eq:nonlineqs1d}) is given by
\begin{gather*}
\Delta x\td{\bm{u}_h}{t} + \LRp{\LRp{\bm{I}_{n \times n} \otimes \bm{Q} }\circ \bm{F}}\bm{1} = \bm{0},\\
\bm{F} = \begin{bmatrix}
\bm{F}_{1} & &\\
& \ddots &\\
&& \bm{F}_{n} \\
\end{bmatrix}, \qquad 
\LRp{\bm{F}_{i}}_{jk} = \bm{f}_{i,S}\LRp{\LRp{\bm{u}_h}_j, \LRp{\bm{u}_h}_k}, \qquad \LRp{\bm{u}_h}_j = \begin{bmatrix}\LRp{\bm{u}^1_h}_j & \LRp{\bm{u}^2_h}_j & \ldots & \LRp{\bm{u}^n_h}_j\end{bmatrix}^T.
\label{eq:essys}
\end{gather*}
Here, $\bm{F}$ is a block diagonal matrix where the $i$th block corresponds to the evaluation of the $i$th component of the numerical flux $\bm{f}_{i,S}(\bm{u}_L,\bm{u}_R)$.  From this point on, we drop explicit references to components and the Kronecker product to simplify notation, so that (\ref{eq:es}) applies to both scalar equations and systems.

@@

If $\bm{v}_h^T\bm{K}\bm{u}_h \geq 0$, then the solution satisfies a discrete dissipation of entropy
\begin{equation}
\Delta x \bm{1}^T\td{S(\bm{u}_h)}{t} = -\epsilon\bm{v}_h^T\bm{K}\bm{u}_h \leq 0.
\label{eq:entropybalancefom}
\end{equation}
The proof is similar to the one given in \cite{tadmor2006entropy}.  We rewrite the viscous term as
\[
\bm{v}_h^T\bm{K}\bm{u}_h = \bm{v}_h^T \frac{1}{\Delta x^2}\begin{bmatrix}
\LRp{\bm{u}_h}_1 - \LRp{\bm{u}_h}_2\\ 
%\LRp{\LRp{\bm{u}_h}_2 - \LRp{\bm{u}_h}_1} - \LRp{\LRp{\bm{u}_h}_3 - \LRp{\bm{u}_h}_2}\\
\vdots\\
\LRp{\LRp{\bm{u}_h}_i - \LRp{\bm{u}_h}_{i-1}} - \LRp{\LRp{\bm{u}_h}_{i+1} - \LRp{\bm{u}_h}_{i}}\\
\vdots\\
\LRp{\bm{u}_h}_K- \LRp{\bm{u}_h}_{K-1} 
\end{bmatrix}.
\]
Since $\bm{v}_h$ and $\bm{u}_h$ are related through a differentiable invertible mapping, the mean value theorem implies that
\[
\LRp{\bm{u}_h}_i - \LRp{\bm{u}_h}_{i-1} = \LRp{\pd{\bm{u}}{\bm{v}}}_{i,i-1} \LRp{\LRp{\bm{v}_h}_i - \LRp{\bm{v}_h}_{i-1}}
\]
where $\LRp{\pd{\bm{u}}{\bm{v}}}_{i,i-1}$ is the Jacobian matrix evaluated at some state between $(\bm{u}_h)_i$ and $(\bm{u}_h)_j$.  Since the Jacobian is related to the Hessian of the entropy through $\pdn{S(\bm{u})}{\bm{u}}{2} = \pd{\bm{v}}{\bm{u}} = \LRp{\pd{\bm{u}}{\bm{v}}}^{-1}$, it is positive definite so long as $S(\bm{u})$ is convex.  Substituting these expressions into the viscous terms and using positive definiteness of $\pd{\bm{u}}{\bm{v}}$  yields 
\begin{gather}
\bm{v}_h^T\bm{K}\bm{u}_h = 
%\bm{v}_h^T \frac{1}{\Delta x^2}\begin{bmatrix}
%\LRp{\pd{\bm{u}}{\bm{v}}}_{2,1} \LRp{\LRp{\bm{v}_h}_1 - \LRp{\bm{v}_h}_{2}}\\ 
%\vdots\\
%\LRp{\pd{\bm{u}}{\bm{v}}}_{i,i-1} \LRp{\LRp{\bm{v}_h}_i - \LRp{\bm{v}_h}_{i-1}} - \LRp{\pd{\bm{u}}{\bm{v}}}_{i+1,i} \LRp{\LRp{\bm{v}_h}_{i+1} - \LRp{\bm{v}_h}_{i}}\\
%\vdots\\
%\LRp{\pd{\bm{u}}{\bm{v}}}_{K,K-1} \LRp{\LRp{\bm{v}_h}_K - \LRp{\bm{v}_h}_{K-1}}
%\end{bmatrix}\\
 \frac{1}{\Delta x^2}\sum_{i=1}^{K-1} \LRp{\pd{\bm{u}}{\bm{v}}}_{i+1,i} \LRb{\LRp{\bm{v}_h}_{i} - \LRp{\bm{v}_h}_{i+1}}^2 \geq 0.
 \label{eq:entropydissfom}
\end{gather}

\section{An entropy conservative reduced order model}
\label{sec:3}
We can now formulate a reduced basis approximation on periodic domains using the entropy stable full order models described in the previous section.  Boundary conditions will be addressed later in Section~\ref{sec:6}.  We also note that this reduced order model is not practical, since the cost of evaluating nonlinear terms scales with the size of the full order model.  Later sections will discuss how to reduce the cost of nonlinear evaluations using appropriate hyper-reduction techniques.

@@

We will show that, while this formulation is not entropy conservative, a slight modification recovers semi-discrete entropy conservation.  We motivate this modification by first considering the time derivative term.  Because of the Galerkin projection, we can no longer directly test with the vector of entropy variables, which may not lie  in the span of the reduced basis functions.  However, we can test with an appropriate \textit{projection} of the entropy variables.  Let $\bm{v}_N$ denote coefficients of the projection of the entropy variables

@@

 \bm{v}\LRp{\bm{V}\bm{u}_N}^T \td{\LRp{\bm{V}\bm{u}_N}}{t} = \bm{1}^T\td{S(\bm{V}\bm{u}_N)}{t}.\nonumber

@@

The remainder of the proof requires showing that 

@@

\bm{v}_N^T\bm{V}^T\LRp{\bm{Q}\circ \bm{F}}\bm{1} = \tilde{\bm{v}}^T\LRp{\bm{Q}\circ \bm{F}}\bm{1}  = 0, \qquad \tilde{\bm{v}} = \bm{V}\bm{v}_N = \bm{V}\bm{V}^{\dagger} \bm{v}\LRp{\bm{V}\bm{u}_N},

@@

where we have introduced the grid values of the projected entropy variables as $\tilde{\bm{v}}$.
We can repeat the steps of the proof of Theorem~\ref{thm:ecfom} up to the point when we invoke the entropy conservation condition of (\ref{eq:esflux})

@@

We apply a two-step hyper-reduction procedure which preserves both the zero row sums and skew-symmetry of the hyper-reduced matrix $\bm{Q}_s$.  Rather than directly hyper-reducing the full order matrix $\bm{Q}$, we first construct a compressed ``modal'' matrix, then combine this with a projection operator based on the hyper-reduced points.  In the first step, we construct a compressed intermediate operator by combining  Galerkin projection with the ``expanded basis'' approach of \cite{hernandez2017dimensional}.  Let $\bm{V}_t$ denote a \textit{test} basis through which to extract the action of $\bm{Q}$.  We define the intermediate reduced operator 
\begin{equation}
\hat{\bm{Q}}_{t} = \bm{V}_{t}^T\bm{Q}\bm{V}_{t}.
\label{eq:compressQ}
\end{equation}
Let $\bm{c}$ be some vector in the span of the test basis such that $\bm{c}=\bm{V}_t \hat{\bm{c}}$. Solving the system $\bm{V}_t^T\bm{V}_t \hat{\bm{d}} = \hat{\bm{Q}}_t\hat{\bm{c}}$ yields that
\[
\bm{V}_t^T\bm{V}_t \hat{\bm{d}} = \bm{V}_t^T\bm{Q}\bm{V}_t\hat{\bm{c}} = \bm{V}_t^T\bm{Q}\bm{c} \qquad \Longrightarrow\qquad \bm{V}_t^T \LRp{\bm{V}_t\hat{\bm{d}}-\bm{Q}\bm{c}} = \bm{0}.
\]
Here, $\hat{\bm{d}}$ are the coefficients of the orthogonal projection of $\bm{Q}\bm{c}$ onto the test basis.  Note that $\hat{\bm{Q}}_t$ is skew-symmetric by construction.  Moreover, if the vector of all ones lies in the span of the test basis (e.g., $\bm{1} = \bm{V}_t\bm{e}$ for some coefficient vector $\bm{e}$), then 
\[
\hat{\bm{Q}}_t\bm{e} = \bm{0},
\]
since $\hat{\bm{Q}}_t$ exactly recovers the action of $\bm{Q}\bm{1} = \bm{0}$ and $\bm{0}$ lies within the span of any basis.
